---
title: "Lesson 1"
author: "Samantha Chinn Nicholas Clark"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,results="hide",fig.show="hide")
```

# Contact Information

LTC Nick Clark

nicholas.clark@westpoint.edu

Thayer Hall Room 226B (Back room, just keep going back...)

845-938-0267

# Course Overview

\textit{Applied } Statistics

-Learn Statistics by doing

-Hands on/interactive

-R/Rstudio

-Course Overview

-Graded Events

# Introductions

# Getting started with R studio/knitr

# Berkely Admissions Data

In the early 1970s, the University of California at Berkeley was concerned with possible
discrimination against women in its graduate admissions process. Data about the applicants for
the 1972–73 school year were recorded from several programs, including their sex and whether
or not they were accepted (Bickel & O’Connell, Science, 1975).

```{r}
#install.packages("tidyverse")
library(tidyverse)
berkley.data<-read.table("http://www.isi-stats.com/isi2/data/Berkeley.txt",header=T)
```

Observational Unit: The noun we are taking the measurement of. The applicants for UC Berkely.

\vspace{.2in}

Response Variable: Whether or not they got accepted into UC Berkely (this specific R.V. is a categorical variable)

\vspace{.2in}

Values Response Variable can take on: Yes (Accepted) or No (Not Accepted)

\vspace{.2in}

Visualize

```{r}
ggplot(data=berkley.data,aes(Accepted))+geom_histogram(stat="count")
```

Explore
```{r,include=F}
no.program<-berkley.data %>% select(-Program)
table(no.program)
```
Visualize the table:

```{r}
#install.packages
library(ggmosaic)
ggplot(data=no.program)+
  geom_mosaic(aes(x=product(Accepted,Sex),fill=Accepted))
```
What's your thoughts? 
\newline That there is potentially some gender bias.

\vspace{.2in}

What would the mosaic plot look like if sex was not associated with acceptance? 
\newline Even split within individual gender columns.

\vspace{.2in}

What other variables might help us predict whether or not someone would be accepted? 
\newline The program they applied to, their GPA, socio-economic background.

\vspace{.4in}

```{r}
ggplot(data=berkley.data)+
  geom_mosaic(aes(x=product(Accepted,Sex),fill=Accepted))+
  facet_grid(Program~.)
```

What are your thoughts looking at this plot? 
\newline More females were accpeted when looking at program A and program F independently.

\vspace{.2in}

How can it be that females have higher acceptance rates than males in Program A AND in Program F but when we combine the two, the overall acceptance rate is noticeably smaller for females? 
\newline Simpson's Paradox - when a trend appears in several different groups of data, but disappears or reverses when the groups are combined.

\vspace{.2in}

In examining gender and acceptance what is \textbf{confounding?} 
\newline Program

\vspace{.2in}

What else could contribute to an admissions decision other than sex and program? 
\newline GPA, socio-economic, geographic

\vspace{1.in}

Sources of Variation Diagram: 
\newline (Example p. 11 of textbook)

\vspace{1.in}

Six Steps of a statistical investigation

\begin{itemize}
\item Ask a Research Question

\item Design a study and collect data

\item Explore the data

\item Draw inferences beyond the data

\item Formulate conclusions

\item Look back and ahead

\end{itemize}


Did we do this with the Berkley Data?  What would you tell the college administrators? 
\newline There may be a problem. There is an impact due to gender, but it could be that gender is influencing program and program is influencing acceptance. Do more investigating: why are fewer females applying to program A than males? Is there a problem in recruiting for program A, is our outreach not as good for program A, is program A really hard and we are discouraging them from applying?

\vspace{.5in}


Let's brainstorm how to write a statistical model for the Berkley data. 
\newline Use the Sources of Variation diagram to help write the statistical model. Some response which is a function of things we explain and we cannot explain. Cannot write a linear regression model because LHS is binary.

\vspace{3.in}

© 2020 GitHub, Inc.
Terms
Privacy
Security
Status
Help
Contact GitHub
Pricing
API
Training
Blog
About
