---
title: "Lesson 2"
author: "Samantha Chinn Nicholas Clark"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.show="hide")
```

In the United States, the 1963 Equal Pay Act requires that men and women be given equal pay for equal work and Title VII of the Civil Rights Act of 1964 prohibits discrimination on the basis of race, color, relgion, sex, and national origin.  How successful have these acts been?

WageRace contains observations from 1987 for a sample of 25,632 males between the age of 18 and 70 who worked full-time along with their years of educaiton, years of experience, race, whether they worked in a standard metropolitan area, and the region of US where they worked.

Primary research question is whether wages for blacks differ significantly from wages for non-blacks?

```{r}
library(tidyverse)
wage.dat<-read.table("http://www.isi-stats.com/isi2/data/Wages.txt",header=T)
```

Identify the observational units in the study.  How many are there?
\newline Males between the age of 18 and 70 who worked full time. 25,631.

```{r}
# nrow counts number of rows in data frame
nrow(wage.dat)
#head(wage.dat) This gives the first couple of entries
```

Is the wages variable a quantitative or categorical variable?
\newline Quantitative

\vspace{.2in}

```{r}
ggplot(wage.dat,aes(x=wage))+geom_histogram(bins=100)
ggplot(wage.dat,aes(y=wage))+geom_boxplot()+coord_flip()
```

Why are we looking at histograms and boxplots rater than a bar graph?
\newline Histograms and boxplots are for quantitative response variables.

\vspace{.2in}

Does anything stand out to you about the boxplot that is less obvious in the histogram?
\newline There are a lot of outliers.

\vspace{.2in}

Which visual, the histogram or boxplot, do you like better?  Why?
\newline I like the boxplot better becuase it shows the outliers.

\vspace{.2in}

Which is larger, the mean or the median?  How do you know?
\newline The mean is larger because it is slightly right skewed, positively skewed. (Median is robust to huge outliers)

\vspace{.2in}

Do the wages appear to follow a normal distribution?  How do you know?
\newline No, because of the positive skewness. With skewness, it is hard to apply the CLT to get a mean.

\vspace{.2in}

In this study, the researchers were most interested in whether race explained differences in wages.

Which variable is the explanatory variable?  Which is the response variable?
\newline Explanatory Variable: race - categorical
\newline Response Variable: wages - quantitative

\vspace{.2in}

Do you think the explanatory variable explains some variation in the response variable?  Do you think it explains all of the variation in the response variable?  Why or why not?
\newline Perhaps it explains some variation, but not all. There may be confounding variables or just other variables. Education, socio-economic, geographic, the job.

\vspace{.2in}
```{r}
ggplot(wage.dat,aes(y=wage,x=race))+geom_boxplot()+coord_flip()
wage.dat%>%group_by(race)%>%
  summarise(n=n(),mean=mean(wage),StDev=sd(wage),Minimum=min(wage),
            Median=median(wage),Maximum=max(wage))
```

Consider whether there appears to be an association between wage and race:  Does the wage distribution differ substantially between blacks and non-blacks?  What is the difference in the mean weekly wages?  Can we conclude wage discrimination?
\newline Yes it appears there is a substantial difference between wages. 654 - 479 = 175. We cannot conclude wage discrimination because we have not yet considered other variables.

\vspace{.2in}
```{r}
ggplot(wage.dat,aes(y=wage,x=educ))+geom_boxplot()+coord_flip()
wage.dat%>%group_by(educ)%>%
  summarise(n=n(),mean=mean(wage),StDev=sd(wage),Minimum=min(wage),
            Median=median(wage),Maximum=max(wage))
```

Suggest an easy way to improve this graphical display to better focus on a trend of increasing salaries with increasing education.
\newline Organize by increasing educational level. This can be fixed in R. Put factors in correct order.

\vspace{.2in}

Describe the association between education and wage.  Is it as you would have predicted?  Explain.
\newline As education goes up, wage increases. This is what I would have predicted, it just makes sense.

\vspace{.2in}

What would need to be true for education level to provide an alternative explanation for why non-blacks in this sample tended to earn more than blacks?
\newline In order for it to be confounding, we would need education to have a relationshiop with both race AND wages.

```{r}
ggplot(wage.dat,aes(y=wage,x=educ,fill=race))+geom_boxplot()+coord_flip()
wage.dat%>%group_by(educ,race)%>%
  summarise(mean=mean(wage),StDev=sd(wage))
```

Is there a difference in the average wage between blacks and non-blacks in the "beyond college" group?  Is this difference larger or smaller than when we did not take the education level into account?
\newline Within each education level non-black has a higher wage, however that difference is smaller now. 971 - 847 = 24.
\newline Simpson's Paradox - as we stratify our groups the difference is smaller.

\vspace{.2in}

Do the lower average wages for blacks compared to non-blacks appear to be consistent across each of the education levels?
\newline Yes it seems consistent.

\vspace{.2in}

If you were to compare the average weekly wage for blacks to the average weekly wage for non-blacks in the same education group, roughly how large would you say that difference is?
\newline Somewhere around 100.

\vspace{.2in}

How do you respond to the argument that the wage disparity between blacks and non-blacks is really an issue of education level?
\newline Yes education level impacts wages a bit, but even withint every single education level, there is still a difference.

\vspace{.2in}

Sources of variation diagram:

\vspace{1.in}

```{r}
birthwt.dat<-read.csv("https://raw.githubusercontent.com/nick3703/MA376/master/births.csv")
```

Explore:

```{r}
ggplot(aes(x=weight),data=birthwt.dat)+geom_histogram(bins=100)
```

Filter out the unknowns

```{r}
birthwt.clean<-birthwt.dat %>% filter(weight < 8166)
ggplot(aes(x=weight),data=birthwt.clean)+geom_histogram(bins=100)
```

summary statistics
```{r}
birthwt.clean%>%summarise(N=n(),Mean=mean(weight),StDev=sd(weight),Min=min(weight),
                          Max=max(weight))
```

If we used the mean to predict future newborn weight how well would we do?
\newline Potentially pretty bad.The Null Model is a bad way of predicting values.

\vspace{.2in}

The statistical model would be:

\vspace{.5in}

A residual is the value $y_i - \hat{y}_i$ for $i =1,\cdots,n$.  We can find the residuals two different ways:

```{r}
birthwt.resid<-birthwt.clean%>% mutate(resid=weight-mean(weight))%>%select(resid)
ggplot(aes(x=resid),data=birthwt.resid)+geom_histogram(bins=100)
birthwt.resid%>%summarise(Mean=mean(resid),StdDev=sd(resid))
```

What is going on?  Have we explained any variation?

\vspace{.5in}

```{r}
ggplot(aes(x=weight,color=full.term),data=birthwt.clean)+geom_histogram(fill="white", 
                                                      alpha=0.5, position="identity")
birthwt.clean%>%group_by(full.term)%>%
  summarise(N=n(),Mean=mean(weight),StDev=sd(weight),Min=min(weight),Max=max(weight))
```


Our predicted model becomes:

\vspace{.5in}

Standard error:
```{r}
model<-lm(weight~0+full.term,data=birthwt.clean)
summary(model)
```

Does Mom's BMI impact weight?

```{r}
birthwt.bmi<-birthwt.clean%>%filter(mom.BMI < 90)
model<-lm(weight~0+full.term*mom.BMI,data=birthwt.bmi)
summary(model) #I think there's an error in book
```

Sources of variation diagram:

\vspace{2.in}

## Memorizing Levels

1-20

11 - Consider histogram, mean, standard deviation, median

13 - Consider side by side boxplots

\vspace{2.in}
© 2020 GitHub, Inc.